---
title: "CRISPRviewR-vignette"
author: Albert C. Vill
date: 21 September 2022
website: https://github.com/acvill/CRISPRviewR
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CRISPRviewR-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = 'center'
)
```

```{r setup}
library(CRISPRviewR)
```

## Read in and merge minCED data

Process file pairs and merge the objects into a single dataset.
```{r, eval=FALSE, message=FALSE, warning=FALSE}
baseurl <- "https://raw.githubusercontent.com/acvill/CRISPRviewR/master/example_data_minced/"

merged <- merge_minced(
  s1 = read_minced(txt = url(paste0(baseurl, "s1.txt")),
                   gff = url(paste0(baseurl, "s1.gff"))),
  s2 = read_minced(txt = url(paste0(baseurl, "s2.txt")),
                   gff = url(paste0(baseurl, "s2.gff"))),
  s3 = read_minced(txt = url(paste0(baseurl, "s3.txt")),
                   gff = url(paste0(baseurl, "s3.gff"))),
  s4 = read_minced(txt = url(paste0(baseurl, "s4.txt")),
                   gff = url(paste0(baseurl, "s4.gff"))),
  s5 = read_minced(txt = url(paste0(baseurl, "s5.txt")),
                   gff = url(paste0(baseurl, "s5.gff")))
  )
```
```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
basedir <- "/Users/acvill/Dropbox/Github/CRISPRviewR/example_data_minced/"

merged <- merge_minced(
  s1 = read_minced(txt = paste0(basedir, "s1.txt"),
                   gff = paste0(basedir, "s1.gff")),
  s2 = read_minced(txt = paste0(basedir, "s2.txt"),
                   gff = paste0(basedir, "s2.gff")),
  s3 = read_minced(txt = paste0(basedir, "s3.txt"),
                   gff = paste0(basedir, "s3.gff")),
  s4 = read_minced(txt = paste0(basedir, "s4.txt"),
                   gff = paste0(basedir, "s4.gff")),
  s5 = read_minced(txt = paste0(basedir, "s5.txt"),
                   gff = paste0(basedir, "s5.gff"))
  )
```

Inspect a subset of the data. Note that running `crispr_summary()` on the full
`merged` object may give a spurious output as `array` labels are repeated 
across samples.
```{r, eval=TRUE, message=FALSE, warning=FALSE}
crispr_summary(merged |> dplyr::filter(id == "s1"))
```

## Group CRISPR arrays by shared repeats

The `group_arrays()` function determines a consensus repeat (`repcon``) for each array,
then associates arrays with identical consensus sequences.
```{r, eval=TRUE, message=FALSE, warning=FALSE}
groups <- group_arrays(merged)
group_summary(groups)
```

Groups that are represented across all of the samples may be interesting to plot. Let's get the labels for those groups.
```{r, eval=TRUE, message=FALSE, warning=FALSE}
group_summary(groups) |>
  dplyr::filter(sample_count == length(unique(merged$id))) |>
  dplyr::select(repcon_grp) |>
  as.vector()
```

## Plot CRISPR arrays

Given a `repcon_grp` number selected from `group_summary()`,
you can subset your grouped data to plot arrays with a shared repeat.
```{r, eval=TRUE, message=FALSE, warning=FALSE}
plot_arrays(group = dplyr::filter(groups, repcon_grp == 116))
```

Now we see the utility of this package! By comparing CRISPR arrays with shared
repeat sequences, we observe that an array resident in sample `s3` lacks the
spacer diversity of the arrays seen in the other samples. **Before interpreting 
these plots as representative of biological truth, be sure to review the [list
of caveats](https://github.com/acvill/CRISPRviewR#caveat-emptor).**

By default, a sequence logo representing the repeat consensus is plotted below
the arrays. This can be toggled with the `plot_logo` switch.
```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.height=3}
plot_arrays(group = dplyr::filter(groups, repcon_grp == 116),
            plot_logo = FALSE)
```



To better differentiate spacers, cluster numbers can be added with the
`number_spacers` option.
```{r, eval=TRUE, message=FALSE, warning=FALSE}
plot_arrays(group = dplyr::filter(groups, repcon_grp == 116),
            number_spacers = TRUE)
```

Spacer clustering is controlled by the `cdist` parameter, which represents a
proportion of the maximum pairwise [Levenshtein distance](https://en.wikipedia.org/wiki/Levenshtein_distance)
for the set of spacers associated with a consensus repeat sequence. The default
value is 0.1. Decreasing this value creates more spacer clusters, and increasing
this value reduces the number of clusters.
```{r, eval=TRUE, message=FALSE, warning=FALSE}
plot_arrays(group = dplyr::filter(groups, repcon_grp == 116),
            cdist = 0.5, number_spacers = TRUE)
```

The default palette used by `plot_arrays()` has 34 colors. Custom palettes
can be specified with the `palette` parameter. If a palette has fewer colors
than there are spacer clusters in the array group, colors are recycled, and 
`number_spacers = TRUE` should be specified.
```{r, eval=TRUE, message=FALSE, warning=FALSE}
alt_palette <- c("#7FC97F","#BEAED4","#FDC086","#FFFF99",
                 "#386CB0","#F0027F","#BF5B17","#666666")
plot_arrays(group = dplyr::filter(groups, repcon_grp == 116),
            palette = alt_palette, number_spacers = TRUE)
```

When a single sample has more than one array with the same consensus repeat
sequence, as in the case when one CRISPR array is split across multiple contigs,
the arrays are plotted side-by-side with gaps.

```{r, eval=TRUE, message=FALSE, warning=FALSE}
plot_arrays(group = dplyr::filter(groups, repcon_grp == 268))
```

